---
title: "Converting gen3sis2 raster spaces to h3 and points"
date: "`r format(Sys.time(),  '%d.%m.%Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting gen3sis2 raster spaces to h3 and points}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  tidy=T,
  fig.align='center',
  tidy.opts = list(width.cutoff=80),
  results='hold'
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In spatial and computational ecology, rasters represent the most common way to handle spatial data. Ecologists commonly use them with environmental variables structured in multiple layers in a uniform square-cell grid, in which each cell contains a unique value for each layer (variable) and represents some area of Earth's surface. Since raster cells do not necessarily cover all the same surface area, rasters suffer greatly from projection distortions, which can be a major source of bias and uncertainty in eco-evolutionary studies. Addressing the distortion issue, many strategies were proposed. For example, H3 system partitions the Earth's surface into multiple hexagonal cells covering the same surface area; thus, information is less biased by projection distortion. Another possible strategy is the use of equidistant points on the planet surface, which also reduces projection bias.

In *gen3sis2*, it is possible to create input spaces from rasters, using the *gen3sis2::create_space_raster* function. These spaces can be later converted to h3 and points using the functions *space_raster_to_h3* and *space_h3_to_raster*, from *spac3tools*. *gen3sis2* is completely compatible with both h3 and points input spaces. In this vignette we will demonstrate a raster spaces and its conversion to h3 and points spaces.  

# Converting space

First, we need to create a input space. To do that, we will use the temperature rasters of South America available at the [simulation repository](https://github.com/project-gen3sis/Simulations). Download it and define the correct path to it in order to continue. We first load the three corresponding raster from the hard drive.

```{r, eval=FALSE, warning=FALSE}
library(terra)

temperature <- rast("~/your_path/SouthAmerica/temperature_rasters.grd")

# set the right crs
crs(temperature) <- "EPSG:4326"

# organize in a raster list
raster_list <- list(
  temperature = temperature
)
```

Now we can use this to create the input space with *gen3sis2::create_spaces_raster*. We will use a simple cost function. 

```{r, eval=FALSE, echo=FALSE, warning=FALSE}
library(gen3sis2)

spaces_output <- "~/your_path/SouthAmerica/temp_spaces"
gen3sis2::create_spaces_raster(
  raster_list = raster_list,
  cost_function = function(source, dest) {
    if(!all(source$habitable, dest$habitable)) {
      return(2/1000)
    } else {
      return(1/1000)
    }
  },
  directions = 8,
  output_directory = spaces_output,
  full_dists = T,
  overwrite_output = T,
  verbose = F,
  duration = list(from = 65, to = 0, by = -1, unit = "Ma"),
  geodynamic = T
)
```

```{r, echo=FALSE, include=FALSE, eval=FALSE}
input_spaces <- "~/your_path/SouthAmerica/temp_spaces/"
config_file <- "~/your_path/SouthAmerica/config/config_southamerica.R"
output_dir <- "~/your_path/SouthAmerica/outputs/"

emulate_space_in_simulation <- function(config, space, output_directory) {
    directories <- gen3sis2::prepare_directories(config_file = config,
                                       input_directory = space,
                                       output_directory = output_directory)
    config <- gen3sis2::create_input_config(config_file = config)
    config[["directories"]] <- directories
    val <- list("data" = list(),
                "vars" = list(),
                "config" = config)

    val$config <- gen3sis2:::complete_config(val$config)
    val$config$gen3sis$general$verbose <- 1
    val <- gen3sis2:::setup_inputs(val$config, val$data, val$vars)
    val <- gen3sis2:::setup_variables(val$config, val$data, val$vars)
    val <- gen3sis2:::setup_landscape(val$config, val$data, val$vars)
    return(val)
}

val <- emulate_space_in_simulation(config_file, input_spaces, output_dir)

gen3sis2::plot_landscape(val$data$landscape)
```

## Converting raster spaces to H3

To convert this raster space to H3 cells, we will use the *space_raster_to_h3* function. All we need is to specify the input and output directories, whether is wanted to compute the distances and the desired H3 resolution. The later is very important, because "resolution" meaning for rasters and H3 are not the same. For example, our raster space resolution is (1,1), or simply 1 degree. This means that each cell covers 1 longitudinal degree in x axis and 1 latitudinal degree in y axis. This information can be retrieved in the spaces.rds itself, in the *$meta$type_spec$res*. Here, *type_spec$res* can be read as "type specific resolution":

```{r, eval=FALSE}
raster_space <- readRDS(file.path(spaces_output,"spaces.rds"))
raster_space$meta$type_spec$res
```

For H3 cells, resolution works differently. H3 works is an hierarchical system, and lower resolutions present cells that are "father" to others. E.g., each hexagon in resolution 0 contains 7 children hexagons of the resolution 1, and each of those have 7 children of the resolution 2, and so on...

For that reason, when converting raster spaces to h3, the resolution is one important decision to take, and should be pondered.

The conversion with *space_raster_to_h3* is very straightforward. The function projects the raster data in any H3 resolution. In the process, values aggregation (e.g., values mean) is often necessary. If "compute_distances" is TRUE, the original distance matrices are updated. If more than one raster cell is inside the same H3 cell, the closest raster cell to the H3 centroid is choosen as the representative. Then, the distance between two H3 cells is assumed as the distance calculated between their representative raster cells.

```{r,eval=FALSE}
library(spac3tools)

h3_spaces_output <- "~/your_path/SouthAmerica/h3"

space_raster_to_h3(dir_input = spaces_output,
                   dir_output = h3_spaces_output,
                   res=3,
                   verbose=0,
                   compute_distances=TRUE,
                   reports=TRUE)
```

## Convert h3 spaces to points

The conversion from type H3 to type Points is done simply changing the type marker in the spaces.rds, what allows *gen3sis2* to proper handle the spaces. It can be done locally ("inplace = TRUE"), overwriting the type H3 spaces.rds, or not ("inplace = FALSE"). If not locally, the function creates and copies all necessary file to a new directory (dir_output) prior conversion.

```{r,eval=FALSE,warning=FALSE}
# Conversion in a new directory
points_spaces_output <- "~/your_path/SouthAmerica/points"
space_h3_to_points(dir_input = h3_spaces_output,
                   dir_output = points_spaces_output,
                   inplace = FALSE)

# Conversion is local
space_h3_to_points(dir_input = h3_spaces_output)
```

# Conclusion

This vignette showed how to convert spaces type raster into type H3 and points  using *spac3tools*. These conversions reduce spatial distortions common in raster grids, offering more consistent and scalable spatial structures for *gen3sis2* simulations. With support for multiple formats, users can choose the most suitable space type for their modelling needs.
