---
title: "Creating a spaces.rds with H3 data"
date: "`r format(Sys.time(),  '%d.%m.%Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a spaces.rds with H3 data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

In gen3sis2, spaces.rds can be created using the `gen3sis2::create_spaces_raster` function, from a named list of rasters. This creates a `raster` type input spaces that can be used within the `gen3sis2::run_simulation` function. However, `gen3sis2` support not only `raster` type inputs, but also `h3` and `points`. `spac3tools` can both convert `raster` type spaces to `h3`, via `space_raster_to_h3` function, and create a`h3`spaces from scratch, using`create_spaces_h3`function, with user-defined H3 addresses. In this vignette, we will see two methods to create an`h3\` type spaces and run a simulation to test our inputs.

# 1. Organize your data

## 1.1 Using `data.frames`

To be able to create an `h3` type spaces, users must have their data organized in a `h3_list`, which consists of a name list of `data.frames`. Each element of the the list must represent one variable. In each `data.frame`, three columns are mandatory:

-   "h3_address": a set of H3 cells addresses
-   "x": the x coordinate of the cell's centroid
-   "y": the y coordinate of the cell's centroid

Obtaining an H3 cells' centroids is easy using the `h3jsr` package, all is needed is a vector of cells addresses:

```{r}
library(h3jsr)
library(spac3tools)

h3jsr::cell_to_point(c("8166fffffffffff", "81663ffffffffff"), simple = F)
```

All of the other columns in each `data.frame` must represent the value of the respective variable at a given timestep. Below is an example of how a `h3_list` should look like:

```{r}
data("h3_list")
head(h3_list$temperature)
```

## 1.2 Using rasters

Most environmental data is available in rasters. For that reason, to create an `h3_list`, users probably will need to:

-   

    1)  Decide which are the H3 cells of interest

-   

    2)  Extract values for this cells from rasters

-   

    3)  Organize the data in a named list

To ease this process, `spac3tools` features the function `data_raster_to_h3`, which transform a list of rasters in a list of data.frames to be used within `create_spaces_h3`. The function takes as inputs:

-   `h3_address`: a set of H3 cells' address
-   `res`: an H3 resolution
-   `raster_list`: a named list of rasters in the same format as used in `gen3sis2::create_spaces_raster`

Users must provide `h3_address` *or* `res`. In case of a set of H3 addresses is passed, the function will extract data from the rasters at the centroid of the cells. If a resolution is provided, the function will try to extract data for *all global cells at the given resolution*. The `raster_list` follows the same format as the one used in `gen3sis::create_spaces_raster`.

Let's follow an example. First, load some raster data:

```{r}
library(terra)

## This is a temperature (Chelsa v2 bio1) raster for South America
## Each layer must be one timestep, and they must be ordered 
temperature_raster <- terra::rast(system.file("extdata/rasters/temperature_rasters_1dg.tif",package = "spac3tools"))

plot(temperature_raster[[62:66]])
```

Now, we can organize it in a `raster_list`:

```{r}
# Names are important, as they denote which variable the raster represents
raster_list <- list(
  temperature = temperature_raster[[62:66]]
)
raster_list
```

Now, will use a set of H3 addresses at resolution 1 that encompasses the South America continent, and create a concice `h3_list`:

```{r}
south_america_cells <- c(
  "8166fffffffffff", "81663ffffffffff", "81677ffffffffff", "815f7ffffffffff", 
  "81667ffffffffff", "818f7ffffffffff", "818a7ffffffffff", "818afffffffffff", 
  "818b7ffffffffff", "818a3ffffffffff", "815f3ffffffffff", "818abffffffffff", 
  "81807ffffffffff", "81803ffffffffff", "8181bffffffffff", "81813ffffffffff", 
  "81a8bffffffffff", "81817ffffffffff", "818bbffffffffff", "81a8fffffffffff", 
  "818b3ffffffffff", "81a87ffffffffff", "81a83ffffffffff", "81a93ffffffffff", 
  "81c2fffffffffff", "81c23ffffffffff", "81c37ffffffffff", "81b2fffffffffff", 
  "81cebffffffffff", "81b27ffffffffff", "81a97ffffffffff", "81b37ffffffffff", 
  "81b23ffffffffff", "81b33ffffffffff", "818e7ffffffffff", "81c33ffffffffff", 
  "81cfbffffffffff", "81df7ffffffffff", "818f3ffffffffff", "8166bffffffffff")

h3_list <- data_raster_to_h3(h3_address = south_america_cells, raster_list = raster_list)

head(h3_list$temperature)
```

It is also possible to provide just an H3 resolution. In this case, the function will try to extract data for all global cells, but will use only the ones within rasters' extent:

```{r}
h3_list_res <- data_raster_to_h3(res = 2, raster_list = raster_list)

head(h3_list_res$temperature)
```

# 2. Creating the spaces.rds

With data properly organized, create an `h3` type spaces follows the same process as in `gen3sis2::create_space_raster`:

```{r}
create_spaces_h3(
    h3_list = h3_list_res,
    cost_function = function(source, dest){
      return(1/1000)
    },
    output_directory = file.path(tempdir(),"h3"),
    full_dists = TRUE,
    overwrite_output = TRUE,
    verbose = TRUE,
    duration=list(from=-4, to=0, by=1, unit="Ma"),
    geodynamic = TRUE
)
space <- readRDS(file.path(tempdir(),"h3","spaces.rds"))
gen3sis2::plot_space_overview(space)
```

Let's compare with the raster version of this spaces:

```{r}
gen3sis2::create_spaces_raster(
    raster_list = raster_list,
    cost_function = function(source, dest){
      return(1/1000)
    },
    output_directory = file.path(tempdir(),"raster"),
    full_dists = TRUE,
    overwrite_output = TRUE,
    verbose = TRUE,
    duration=list(from=-4, to=0, by=1, unit="Ma"),
    geodynamic = TRUE
)
space_r <- readRDS(file.path(tempdir(),"raster","spaces.rds"))
gen3sis2::plot_space_overview(space_r)
```

# 2.1 Resolution advantages

One advantage of creating an `h3` spaces from scratch instead of converting a `raster` one is the resolution freedom. For example, if we try to convert the raster spaces we created to an H3 resolution of 2, it will not end well, because the area that each raster cells cover at 5 degrees resolution is much different of the area each H3 cell cover at resolution 3, causing blank spaces between cells to appear:

```{r}
space_raster_to_h3(
  file.path(tempdir(),"raster"),
  file.path(tempdir(),"h3_converted"),
  res = 2
)
space <- readRDS(file.path(tempdir(),"h3_converted","spaces.rds"))
gen3sis2::plot_space_overview(space)
```

Now, if we create an `h3` spaces using the same `raster_list`, the results will be much better:

```{r}
h3_list <- data_raster_to_h3(res = 2, raster_list = raster_list)

create_spaces_h3(
    h3_list = h3_list,
    cost_function = function(source, dest){
      return(1/1000)
    },
    output_directory = file.path(tempdir(),"h3"),
    full_dists = TRUE,
    overwrite_output = TRUE,
    verbose = TRUE,
    duration=list(from=-4, to=0, by=1, unit="Ma"),
    geodynamic = TRUE
)
space <- readRDS(file.path(tempdir(),"h3","spaces.rds"))
gen3sis2::plot_space_overview(space)
```

# Running a simulation
